\subsection{From a transducer semigroup to a regular function}
\label{sec:hard}
We now turn to the difficult implication $(\ref{it:trans-semig-f1}) \Rightarrow (\ref{it:regular})$ in Theorem~\ref{thm:regular-functions}. 
 The proof is presented in a way which, if sometimes slightly verbose, makes it
 easier to see how it can be adapted to other algebraic structures instead of
 semigroups (such as forest algebras, cf.~\Cref{sec:conclusion}).

\subsubsection{\kl{Functorial streaming string transducers}}
\label{sec:abstract-sst} 
The assumption of the implication uses an abstract model (\kl{transducer semigroups}), while the conclusion uses a concrete operational model (\kl{streaming string transducers}). To bridge the gap, we use an intermediate model, similar to \sst{}s, but a bit more abstract. The abstraction arises by using \kl{polynomial functors} instead of registers, as described below. 

\AP Define a \intro{polynomial functor} to be a semigroup-to-set functor of the form
\begin{align*}
\qquad A \quad \mapsto \quad \coprod_{q \in Q} A^{\text{\kl{dimension} of } q},
\end{align*}
where $Q$ is some possibly infinite set, whose elements are called \intro{components}, with each \kl{component} having an associated \intro{dimension} in $\set{0,1,\ldots}$. The symbol $\coprod$ stands for disjoint union of sets. This functor does not take into account the semigroup structure of the input semigroup, since the output is seen only as a set.
On morphisms, the functor works in the expected way, i.e.~coordinate-wise.  

\AP A \intro{finite polynomial functor} is a \kl{polynomial functor} with finitely many \kl{components} -- for example, $A \mapsto A^2 + A^2 + A$. 
The notion of \kl{finite polynomial functor} can be seen as a mild generalization of the construction which maps a semigroup $A$ to the set $A^R$ of register valuations for some fixed finite set $R$ of register names.  In the generalization, we allow a variable number of registers, depending on some finite information (the component). 

Having defined a more abstract notion of \enquote{register valuations}, namely \kl{finite polynomial functors}, we now define a more abstract notion of \enquote{register updates}. The first condition for such updates is that they do not look inside the register contents; this condition is captured by \kl{naturality} as described in the following definition. 



\begin{definition}[\intro{Natural functions}]\label{def:natural-functions}
    Let $\functor$  and $\functorg$ be polynomial functors, let $A$ be a semigroup. A function\footnote{This function is not necessarily a semigroup homomorphism. In fact, it would not even make sense call it a homomorphism, since the functors $\functor$ and $\functorg$ produce sets and not semigroups.}  $f : \functor A \to \functorg A$ is called \emph{natural} if it can be extended to a \kl{natural transformation} of type $\functor \Rightarrow \functorg$. This means that there is a family of functions, with one function
    $
    f_A : \functor A \to \functorg A
    $
    for each semigroup $A$, such  that $f=f_A$, and the the  diagram
    \[
    \begin{tikzcd}
    \functor A 
    \ar[r,"\functor h"]
    \ar[d,"f_A"']
    & 
    \functor B 
    \ar[d,"f_B"]
    \\
    \functorg A 
    \ar[r,"h"]
    &
    \functorg B
    \end{tikzcd}
    \]
    commutes for every semigroup homomorphism $h$.
\end{definition}

% Intuitively speaking, naturality says that the function is not allowed to look inside the semigroup elements that are stored in a polynomial functor $\functor A$, but it is allowed to manipulate them using the semigroup operation.

\begin{example}
    Consider the \kl{polynomial functors} (where $1$ represents the singleton set $A^0$)
    \begin{align*}
    \functor A = A^* = 1 + A^1 + A^2 + \cdots  \qquad \functorg A = A + 1.     \end{align*} 
An example of a \kl{natural transformation} between these two functors is the function which maps a nonempty list in $A^*$ to the product of its elements, and which maps the empty list to the unique element of $1$. A non-example is the function that maps a list $[a_1,\ldots,a_n] \in A^*$ to the leftmost element $a_i$ that is an idempotent in the semigroup, and returns $1$ if such an element does not exist. The reason why the non-example is not natural is that a semigroup homomorphism can map a non-idempotent to an idempotent.
\end{example}
% The naturality condition is an abstraction of the condition (a) that was mentioned before, namely that register contents are not inspected. For example, as we will see below, a natural transformations from the functor $\functor A = A^2$ to itself is $(a,b) \mapsto (abbaa,bbba)$, while 
% \begin{align*}
% (a,b) \mapsto \begin{cases}
%     (1,1) & \text{if the semigroup $A$ has an identity element $1$, and $a=1$}\\
%     (a,b) & \text{otherwise.}
% \end{cases}
% \end{align*}
% is not a natural transformation.


Apart from \kl{naturality}, we will want our register updates to be copyless. 

\begin{definition}[\intro{Copyless natural function}] \label{def:copyless} A natural function $f : \functor A \to \functorg A$ is  called \emph{copyless} if it arises from some natural transformation with the following property:  when instantiated to the semigroup\footnote{The choice of the semigroup $(\Nat,+)$ in the \Cref{def:copyless} is not particularly important. For example, the same notion of copylessness would arise if instead of $(\Nat,+)$, we used the semigroup $\set{0,1}$ with addition up to threshold $1$ (i.e.~the only way to get zero is to add two zeros). In the appendix, we present a more syntactic characterization of copyless natural transformations, which will be used later on when proving equivalence with streaming string transducers. 
    } $(\Nat,+)$, the corresponding function of type $\functor \Nat \to \functorg \Nat$ 
    does not increase the norm. Here, the norm of an element in a polynomial functor $\functor \Nat$ or $\functorg \Nat$ is defined to be the sum of numbers that appear in it.
\end{definition}

Having defined functions that are natural and copyless, we now describe the more
abstract model of \sst{}s used in our proof. The main difference is that instead
of register valuations and updates given by some finite set of register names,
we have two abstract \kl{polynomial functors}, one of them \kl{finite polynomial},
together with an explicitly given application function. We also allow the
computation to be initialized and finalized in a more liberal way, that may depend on a
regular property of the input.

\begin{definition}\label{def:functorial-sst}
    The syntax of a \intro{functorial streaming string transducer} is given by:
    \begin{itemize}
        \item A finite \emph{input alphabet} $\Sigma$ and an \emph{output semigroup $A$}.
    \item A \kl{finite polynomial functor} $\functorr$, called the \intro{register functor}, and a (not necessarily finite) \kl{polynomial functor} $\functoru$ called the \intro{update functor}.
    \item A \kl{copyless natural function} of type $\functorr A \times \functoru A \to \functorr A$, called \emph{application}.
    \item An \intro{initial function} $\Sigma^* \to \functorr A$ which is \kl{recognizable} (and therefore has finite range).
    \item A polynomial \intro{final data functor} $\functork$, a \intro{final data function} $\Sigma^* \to \functork A$ which is \kl{recognizable}, and a \intro{final output function} of type $\functorr A \times \functork A \to A$ which is a \kl{natural function} (but not necessarily copyless).
    \item An \emph{update oracle}, which is a \kl{rational function} of type $\Sigma^* \to (\functoru A)^*$.
    \end{itemize}
\end{definition}
The semantics of the \functorialsst is a function 
$\Sigma^* \to A$ defined as follows. Given an input string, we feed it to the
\kl{initial function} to get an initial register valuation. As in
Definition~\ref{def:usual-sst}, we use the application function and the update
oracle to define inductively a sequence of register valuations in $\functorr A$.
We then apply the \kl{final output function} to the last register valuation and to the
final data (determined by the input), yielding a result in $A$.

In the appendix, we prove that this model is no more expressive than usual copyless \sst{}s.

\begin{lemma}\label{lem:functorial-sst-complete}
    Definitions~\ref{def:usual-sst} and~\ref{def:functorial-sst} characterize the same string-to-semigroup functions.
\end{lemma}
