\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate, numberwithinsect]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{A Categorical Perspective on Regular Functions} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Mikołaj Bojańczyk}{Institute of Informatics, University of Warsaw, Poland \and \url{https://www.mimuw.edu.pl/~bojan/}}{bojan@mimuw.edu.pl}{}{{\color{red}(Optional) author-specific funding acknowledgements}}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Lê Thành Dũng (Tito) Nguy\~{\^e}n}{École normale supérieure de Lyon, France \and \url{https://nguyentito.eu/}}{nltd@nguyentito.eu}{https://orcid.org/0000-0002-6900-5577}{Supported by the LABEX MILYON (ANR-10-LABX-0070) of Université de Lyon, within the program \enquote{Investissements d'Avenir} (ANR-11-IDEX-0007) operated by the French National Research Agency (ANR).}

\authorrunning{M.~Bojańczyk and L.~T.~D.~Nguy\~{\^e}n} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Mikołaj Bojańczyk and Lê Thành Dũng Nguy\~{\^e}n} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Theory of computation~Transducers} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm 


\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

\nolinenumbers %uncomment to disable line numbering



%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{csquotes}
\usepackage{mathtools}
\usepackage{quiver}
\usepackage{tikz-cd}
\usepackage{xspace}

\usepackage{todonotes}
\newcommand{\tito}[1]{\todo[inline,color=green!40]{Tito --- #1}}
\usepackage{macros}


\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
    We consider regular string-to-string functions, i.e.~functions that are recognized by copyless streaming string transducers, or  any of their equivalent models, such as deterministic two-way automata. We give yet another characterization, which is very succinct: finiteness-preserving functors from the category of semigroups to itself, together with a certain output function that is a natural transformation.
\end{abstract}

\newcommand{\moncat}{\mathrm{Mon}}
\newcommand{\semcat}{\mathrm{Sem}}

\input{intro}
\input{transducer-semigroup}
\input{output-monoid}



%%
%% Bibliography
%%
%% Please use bibtex, 

\bibliography{bib}

\appendix

\newpage

\tito{Regular functions over arbitrary semigroups!\\
  Instead of just \enquote{transducer semigroups} we're going to prepend the adjective \enquote{functorial}, to distinguish them from the new \enquote{register transducer semigroups}. The notion of regular semigroup-to-semigroup function that we consider is closed under composition (\Cref{rem:composition}) and when the codomain is finite, it coincides with recognizable functions.}

\section{(Attempt at an alternative) Introduction}

The starting point of this paper is the fundamental class of regular
string-to-string functions. It has many equivalent descriptions: {\color{red}\bf
  [\ldots]}

Some generalizations of this class are well understood. For instance, MSO transductions, streaming string transducers\footnote{The suitable generalization of the copyless streaming string transducer is the \emph{single use restricted macro tree transducer}~\cite{MacroMSO} which predates SSTs by two decades while being very similar.} and the formalisms based on the linear $\lambda$-calculus\footnote{The results of~\cite{LambdaTransducer} applies \enquote{out of the box} to trees. As for the \enquote{implicit automata in typed $\lambda$-calculi} approach~\cite{IATLC,titoPhD}, its extension to trees requires the use of the \emph{additive connectives} of linear logic.} admit extensions to tree-to-tree functions that are still equivalent, leading to a robust notion of regular tree functions for which functional combinators can also be designed~\cite{FOTree}.

The present paper is concerned with a different and new generalization of the
string-to-string case: we propose a notion of regular functions \emph{between
  arbitrary semigroups}, which coincides with the regular string-to-string
functions when the semigroups under consideration are free monoids. Some of the
aforementioned models, such as two-way automata or streaming string transducers,
make sense when the output belongs to an arbitrary semigroup; however, all of
them depend on the string structure of the input. This is why we introduce two
new devices that can operate over any semigroup, which we prove to be equivalent
in expressive power to argue for the robustness of our definition. Their
specialization to free monoids adds two new characterizations of the regular
string-to-string functions to the above list.
\begin{description}
\item[Register transducer semigroups] are to streaming string transducers what
  semigroups or monoids are to deterministic finite automata. An SST basically consists of a DFA plus some string-valued registers, while a register transducer semigroup consists of a finite control semigroup plus some registers with values in the output semigroup. A significant difference made by replacing the DFA by a semigroup is that we do not need any \enquote{copyless} or \enquote{bounded-copy} condition anymore to capture precisely the regular functions.
    \item[Functorial transducer semigroups] use minimal syntax, and refer only to basic concepts from algebra and category theory. The characterization is concise enough to be fully stated here: a regular function $A\to B$ is one that can be decomposed as
    \[
        \begin{tikzcd}
            A
            \ar[r,"h"]
            & 
            \functor B
            \ar[r,"\outfun_B"]
            &
            B
        \end{tikzcd}
    \]
    where $\functor$ is some finiteness-preserving functor from the category of semigroups to itself, $h$~is some semigroup homomorphism, and the output function $\outfun_{B}$ is a natural transformation. From this category-theoretic definition, it is easy to show that regular semigroup-to-semigroup functions are \emph{closed under composition}.
    \tito{refer to difficulties with 2DFT / SST composition here? (as in \Cref{rem:composition})}
\end{description}

\section{Preliminaries: streaming string transducers}

\tito{Taken from comparison-free paper with Cécilia. We might want to rephrase
  some things to avoid leaking the authors' identity}

\newcommand{\rmst}{\mathrm{st}}
\newcommand{\rmreg}{\mathrm{reg}}
\newcommand{\rmout}{\mathrm{out}}
\newcommand{\deltast}{\delta_\rmst}
\newcommand{\deltareg}{\delta_\rmreg}
\newcommand{\deltaout}{\delta_\rmout}
\newcommand{\ttout}{\mathtt{out}}
\newcommand{\ttreverse}{\mathtt{reverse}}

\begin{definition}
  \label{def:register-substitution}
  Fix a finite alphabet $\Sigma$. Let $R$ and $S$ be two finite sets
  \emph{disjoint from $\Sigma$}; we shall consider their elements to be
  \enquote{register variables}.

  For any word $\omega \in (\Sigma \cup R)^*$, we write
  $\omega^\dagger : (\Sigma^*)^R \to \Sigma^*$ for the map that sends
  $(u_r)_{r \in R}$ to $\omega$ in which every occurrence of a register variable
  $r \in R$ is replaced by $u_r$ -- formally, we apply to $\omega$ the morphism
  $(\Sigma \cup R)^* \to \Sigma^*$ that maps $c \in \Sigma$ to itself and
  $r \in R$ to $u_r$.

  A \emph{register substitution}\footnote{The name can be found in e.g.~\cite{AlurFT12,AperiodicSST}.} $\alpha$ from $R$ to $S$ (over $\Sigma$) is a map
  $\alpha : S \to (\Sigma \cup R)^*$. It induces the action $\alpha^\dagger :
  \vec{u} \in (\Sigma^*)^R \mapsto (\alpha(s)^\dagger(\vec{u}))_{s \in S} \in
  (\Sigma^*)^S$ (which indeed goes \enquote{from $R$ to $S$}).
\end{definition}

\begin{definition}[\cite{CopyfulSST}]
  \label{def:sst}
  A (deterministic copyful) \emph{streaming string transducer} (SST) with input
  alphabet $\Gamma$ and output alphabet $\Sigma$ is a tuple $\mathcal{T} = (Q, q_0, R,
  \delta, \vec{u}_I, F)$ where
\begin{itemize}
\item $Q$ is a finite set of \emph{states} and $q_0 \in Q$ is the \emph{initial
    state};
\item $R$ is a finite set of \emph{register variables}, that we require to be
  \emph{disjoint from $\Sigma$};
\item $\delta : Q \times \Gamma \to Q \times (R \to (\Sigma \cup R)^*)$ is the
  \emph{transition function} -- we abbreviate $\deltast = \pi_1 \circ \delta$
  and $\deltareg = \pi_2 \circ \delta$, where $\pi_i$ is the projection from
  $X_1 \times X_2$ to its $i$-th component $X_i$;
\item $\vec{u}_I \in (\Sigma^*)^R$ describes the \emph{initial register values};
\item $F : Q \to (\Sigma \cup R)^*$ describes how to recombine the final values
  of the registers, depending on the final state, to produce the output.
\end{itemize}
% To elucidate the last item, observe that $(\Sigma \cup R)^*$ is in canonical
% bijection with the set $\{\ttout\} \to (\Sigma \cup R)^*$ of register
% substitutions from $R$ to a singleton set $\{\ttout\}$. This means that for
% all $q \in Q$, $F(q)$ induces a map $F(q)^\dagger : (\Sigma^*)^R \to \Sigma^*$,
% consistently with the $(-)^\dagger$ notation for substitutions (with $\Sigma^*
% \cong (\Sigma^*)^{\{\mathrm{out}\}}$).
% tout ce paragraphe est économisé grâce au choix de notation \omega^\dagger
The function $\Gamma^* \to \Sigma^*$ computed by $\mathcal{T}$ is
\[ w_1 \ldots w_n ~~ \mapsto ~~ F(q_n)^\dagger \circ \deltareg(q_{n-1},
  w_n)^\dagger \circ \ldots \circ \deltareg(q_0, w_1)^\dagger(\vec{u}_I) \]
where the sequence of states $(q_i)_{0 \leq i \leq n}$ (sometimes called the
\emph{run} of the transducer over the input word) is inductively defined,
starting from the fixed initial state $q_0$, by $q_i = \deltast(q_{i-1},w_i)$.
\end{definition}

\begin{definition}[Alur and {\v{C}}ern{\'{y}}~{\cite{alurExpressivenessStreamingString2010}}]
  A register substitution $\alpha : S \to (\Sigma \cup R)^*$ from $R$ to $S$ is
  said to be \emph{copyless} when each $r \in R$ \emph{occurs at most once among
    all the strings $\alpha(s)$ for $s \in S$}, i.e.\ it does not occur at least
  twice in some $\alpha(s)$, nor at least once in $\alpha(s)$ and at least once
  in $\alpha(s')$ for some $s \neq s'$. (This restriction does not apply to the
  letters in $\Sigma$.)
  
  A streaming string transducer is \emph{copyless} if all the substitutions in the
  image of its transition function are copyless. In this paper, we take
  computability by copyless SSTs as the definition of \emph{regular functions}.
\end{definition}


\begin{example}[Iterated reverse]
  \label{ex:iterated-reverse}
  The following single-state SST is copyless:
  \[ \Gamma = \Sigma\ \text{with}\ \# \in \Sigma \qquad Q = \{q\} \qquad R =
    \{X,Y\} \qquad \vec{u}_I = (\varepsilon)_{r \in R} \qquad F(q) = XY\]
  \[ \delta(q,\#) = (q,\, (X \mapsto XY\#,\, Y \mapsto \varepsilon)) \qquad
    \forall c \in \Sigma \setminus \{\#\},\; \delta(q,c) = (q,\, (X \mapsto X,\,
    Y \mapsto cY)) \]
  For $u_1,\ldots,u_n \in (\Sigma \setminus \{\#\})^*$, it maps
  $u_1\#\ldots\#u_n$ to $\ttreverse(u_1)\#\ldots\#\ttreverse(u_n)$.
\end{example}

\tito{Finite coyping~\cite{MacroMSO} = bounded-copy~\cite{AlurFT12} (TODO: check)}

\section{Register transducer semigroups}

\tito{Probably in the final paper it would be nice to introduce these things without invoking categories at first, to make the paper more welcoming}

\begin{definition}
    A \emph{finitary semigroup with registers} consists of a finite semigroup $S$, a family $(R_s)_{s\in S}$ of finite nonempty sets, and a family of functions
    \[ \mu_{s_1, s_2} : R_{s_1 s_2} \to (R_{s_1} + R_{s_2})^+ \quad\text{for}\ s_1, s_2 \in S\]
    such that for all $s_1, s_2, s_3 \in S$, the following associativity diagram
    commutes:
    \[\begin{tikzcd}
        [column sep=2.5cm]
        &
        (R_{s_1 s_2} + R_{s_3})^+
        \ar[r,"(\mu_{s_1,s_2}+\mathrm{id})^+"']
        &
        ((R_{s_1}+R_{s_2})+R_{s_3})^+
        \ar[dd,"\text{assoc.\ of}\ +"]
        \\
        R_{s_1 s_2 s_3}
        \ar[ur,"\mu_{s_1s_2,s_3}"]
        \ar[dr,"\mu_{s_1,s_2s_3}"']
        \\
        &
        (R_{s_1} + R_{s_2 s_3})^+
        \ar[r,"(\mathrm{id}+\mu_{s_2,s_3})^+"]
        &
        (R_{s_1}+(R_{s_2}+R_{s_3}))^+
      \end{tikzcd}\]
    This data induces the finite polynomial functor from semigroups to sets
    \[ A \mapsto \sum_{s\in S} A^{R_s} \]
    which can be turned into a functor from semigroups to semigroups thanks to $\mu$.
\end{definition}

\begin{remark}
    Equivalently, consider the following symmetric monoidal category:
    \begin{description}
        \item[Objects:] finite polynomial functors (in one variable).
        \item[Morphisms:] natural transformations.
        \item[Monoidal product:] pointwise cartesian product of sets.
    \end{description}
    A finitary semigroup with registers is the same thing as a \enquote{semigroup object} (defined by removing the unit requirement from monoid objects) in this monoidal category of finite polynomial functors.
\end{remark}

\begin{definition}
    A \emph{register transducer semigroup} is a finitary semigroup with registers $(S, (R_s)_{s\in S},(\mu_{s_1,s_2})_{s_1,s_2\in S})$, plus a family of strings $\omega_s \in (R_s)^+$ for $s\in S$.
\end{definition}

This is equivalent to specifying a natural transformation from the associated polynomial functor to the forgetful functor from semigroups to sets, so it induces a functorial transducer semigroup in the expected way.

\begin{definition}
    A \emph{regular function} from $A$ to $B$ (both semigroups) is one that is recognized by the functorial transducer semigroup induced by some 
\end{definition}

\begin{theorem}
    When $A$ and $B$ are free monoids, this coincides with the usual notion.
\end{theorem}
\begin{proof}[Proof idea]
    Easy direction: take the substitution transition monoid~\cite{AperiodicSST} of a bounded-copy SST. (This requires some care to handle the output letters produced not by a transition but by the final output function; they can be stored in specific constant registers that are initialized in the image of all words, including the empty word -- for this, it is important that we can use a morphism of semigroups that is not necessarily a morphism of monoids.)
    
    Conversely, from a register transducer semigroup one can define an SST, and thanks to associativity(!) turns out that it's automatically bounded-copy.
\end{proof}

\begin{remark}
This should also work if we have a single finite nonempty set of registers $R$ for all semigroup elements instead of a family $(R_s)_{s\in S}$ though the equivalence proof is slightly less immediate for semigroups than for monoids (you need to use the output function to produce \enquote{default} elements with which to fill unused registers). Note that the \enquote{substitution transition monoid} operation \enquote{naturally} produces something with a varying register set.

Another possible variation on the definition is to have a choice of output register for each element of the control semigroup: $r_s \in R_s$ for $s\in S$ instead of $\omega_s \in (R_s)^+$. This does not decrease the expressivity of the model.
\end{remark}

\section{Equivalence with functorial transducer semigroups}

It is immediate that every regular function is recognized by a finiteness-preserving functorial transducer semigroup, because finite polynomial functors are finiteness-preserving. We shall therefore focus on the converse.

\begin{theorem}
  For every finiteness-preserving functorial transducer semigroup $(\functor, \outfun)$, there is another one $(\functor',\outfun')$
  \begin{itemize}
    \item which is induced by some register transducer semigroup,
    \item and which admits a natural transformation $\eta : \functor \Rightarrow \functor'$ such that $\outfun = \outfun' \circ \eta$.
  \end{itemize}
  Hence every function recognized by $(\functor,\outfun)$ is also recognized by $(\functor',\outfun')$ (by the second item) and thus is regular (by the first item).
\end{theorem}
\begin{proof}
    Let $(\functor,\outfun)$ be a functorial transducer semigroup, and suppose $\functor$ preserves finiteness, so $\functor1$ is finite. We first build a finitary semigroup with registers that doesn't entirely fulfill our objectives, but still contains the core mechanism: $S = \functor1$, and for $s\in S$,
    \[ R_s = \sum_{l,r \in F(1)} \{\text{occurrences of}\ \mathtt{m}\ \text{in}\ (\text{factorized output of}\ (l,s,r)) \in 1\oplus1\oplus1 \cong \{\mathtt{l,m,r}\}^+\} \]
    For $s_1,s_2\in S$ we must now define a map $R_{s_1 s_2} \to (R_{s_1} + R_{s_2})^+$ describing the register recombination during a product. We derive it from the correspondence between
    \begin{itemize}
        \item the occurrences of $\mathtt{m}$ in the factorized output of $l \cdot (s_1 s_2)\cdot r$
        \item and the maximal factors in $\{\mathtt{m_1,m_2}\}^+$ inside
        \[ (\text{factorized output of}\ (l,s_1,s_2,r)) \in 1\oplus1\oplus1\oplus1 \cong \{\mathtt{l,m_1,m_2,r}\}^+ \]
    \end{itemize}
    Indeed, the morphism $1\oplus\mathop{!}\oplus1$ where $! : 1\oplus1 \to 1$ is the terminal morphism sends the maximal factors in $\{\mathtt{m_1,m_2}\}^+$ to $\mathtt{m}$ and is the identity on $\mathtt{l,r}$. To see that applying this to the factorized output of $(l,s_1,s_2,r)$ indeed yields the factorized output of $(l,s_1s_2,r)$, apply \Cref{lem:merge-middle} to $A=B=C=1$.

    We must also check that this operation is \enquote{associative}. For this, consider the factorized output of $(l,s_1,s_2,s_3,r)$ \ldots{}
    \tito{TODO above}

    Finally we define $\widehat\eta : \functor \Rightarrow \functorg$ where $\functorg$ is the functor induced by our semigroup with registers. For any $x \in \functor A$, apply $\functor! : \functor A \to \functor 1$ to get some $s \in S$. Then look at the factorized output of $(l,x,r)$ in $1\oplus A \oplus 1$ for each $(l,r)\in (\functor1)^2$ to get register values in $A^{R_s}$.

    The issue is that the output function for $\functor$ may not factor through $\widehat\eta$. To remediate that, we consider instead another semigroup with registers, inducing the functor $\functor'$, with the same underlying semigroup $S=\functor1$ and bigger register sets
    \[ R'_s = \{\bullet\} + R^{\mathrm{left}}_s + R^{\mathrm{right}}_s + R_s \quad\text{for}\ s\in S\]
    where $R^{\mathrm{left}}_s$ is the sum over $r \in \functor1$ of the sets of occurrences of $\mathtt{l}$ in the factorized output of $(s,r)$ which is in $1\oplus1\cong\{\mathtt{l,r}\}^+$, etc. Thus
    \[ \functor'A \cong \sum_{s\in S} A \times A^{R^{\mathrm{left}}_s} \times \dots \cong A \times \sum_{s\in S} A^{R^{\mathrm{left}}_s} \times \dots \]
    and we have then: $\outfun_A = (\text{left projection on}\ A) \circ \eta$ for the definition of $\eta$ extending $\widehat\eta$ in the obvious way.
\end{proof}

\begin{lemma}\label{lem:merge-middle}
    Let $A,B,C$ be semigroups. The following diagram commutes:
    \[\begin{tikzcd}
        [column sep=3cm]
        \functor A \times \functor B\times \functor B \times \functor C
        \ar[r,"\text{factorized output}"]
        \ar[d,"\functor A \times (\text{semigroup operation}) \times \functor C"']
        &
        A \oplus B \oplus B \oplus C
        \ar[d,"A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C"]
        \\
        \functor A \times \functor B \times \functor C
        \ar[r,"\text{factorized output}"]
        &
        A \oplus B \oplus C
    \end{tikzcd}\]
\end{lemma}
\tito{similar to \Cref{claim:merge-factorized-output}}
\begin{proof}
    By rotating the above diagram, we see that it corresponds to the outer rectangle in the following diagram (where we have expanded the definition of the factorized output functions, and used the associativity of 4-fold multiplication):
    \[\begin{tikzcd}
        [column sep=4cm]
        \functor A \times \functor B\times \functor B \times \functor C
        \ar[d,"\functor(\text{co-projection})^4"']
        \ar[r,"\functor A \times (\text{semigroup op}) \times \functor C"]
        &
        \functor A \times \functor B \times \functor C
        \ar[d,"\functor(\text{co-projection})^3"]
        \\
        \functor(A \oplus B \oplus B \oplus C)^4
        \ar[d,"\text{multiply middle two together}"']
        &
        \functor(A \oplus B \oplus C)^3
        \ar[dd,""]
        \\
        \functor(A \oplus B \oplus B \oplus C)^3
        \ar[d,"\text{semigroup operation}"']
        \ar[ru,"F(A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C)^3"']
        \\
        \functor(A \oplus B \oplus B \oplus C)
        \ar[d, "\outfun_{A \oplus B \oplus B \oplus C}"']
        \ar[r,"F(A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C)"]
        &
        \functor(A \oplus B \oplus C)
        \ar[d, ""]
        \\ 
        A \oplus B \oplus B \oplus C
        \ar[r,"A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C"]
        &
        A \oplus B \oplus C
    \end{tikzcd}\]
    The lower rectangle commutes by naturality of the output function. The middle trapeze commutes because of the homomorphism property of $F(A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C)$. Concerning the upper trapeze, it can be decomposed as the functor $(\cdots \times \cdots \times \cdots)$ applied to three diagrams that we can analyze independently. The first of those is
    \[\begin{tikzcd}
        [column sep=4cm]
        \functor A
        \ar[d,"\functor(\text{co-projection})"']
        \ar[r,"\mathrm{id}_A"]
        &
        \functor A
        \ar[d,"\functor(\text{co-projection})"]
        \\
        \functor(A \oplus B \oplus B \oplus C)
        \ar[d,"\mathrm{id}"']
        &
        \functor(A \oplus B \oplus C)
        \\
        \functor(A \oplus B \oplus B \oplus C)
            \ar[ru,"F(A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C)"']
    \end{tikzcd}\]
    By functoriality of $\functor$, the commutativity of this diagram reduces to
    \[ (\mathrm{id}_A \oplus \dots) \circ (\text{co-projection of}\ A\ \text{into}\ A\oplus(B\oplus B\oplus C)) = \text{co-projection of}\ A\ \text{into}\ A\oplus(B\oplus C) \]
    which is an elementary property of the coproduct $\oplus$ in any category (indeed the bifunctor structure of $\oplus$ is built using a coparing, which \enquote{cancels out} with the coprojection).
    Among the 3 diagrams that were combined by a 3-fold cartesian product, another one is identical to the above, with $C$ replacing $A$. There remains this one, in which we have added some parts in the middle:
    \[\begin{tikzcd}
        [column sep=2.5cm]
        \functor B\times \functor B
        \ar[d,"\functor(\text{co-proj})^2"']
        \ar[dr,"\functor(\text{co-proj})^2"]
        \ar[rr,"\text{semigroup operation}"]
        &
        &
        \functor B
        \ar[d,"\functor(\text{co-proj})"]
        \\
        \functor(A \oplus B \oplus B \oplus C)^2
        \ar[d,"\text{semigroup op}"']
        \ar[r,"h\times h"']
        &
        \functor(A \oplus B \oplus C)^2
        \ar[r,"\text{semigroup op}"]
        &
        \functor(A \oplus B \oplus C)
        \\
        \functor(A \oplus B \oplus B \oplus C)
            \ar[rru,"h = F(A \oplus (\mathrm{id}_B\,\text{or}\,\mathrm{id}_B) \oplus C)"']
    \end{tikzcd}\]
    The upper right trapeze commutes because $\functor$ applied to the co-projection of $B$ into $A\oplus B\oplus C$ is a homomorphism. The lower triangle commutes because $h$ is a homomorphism. Finally, the small upper left triangle can be shown to commute using the functoriality of $\functor$ followed by properties of the coproduct.
\end{proof}


\end{document}
