\section{Transducer semigroups and warm-up theorems}\label{sec:warm-up}
% We use some basic notions from category theory, such as functors or natural transformations. 
We do not assume that the reader has a background in category theory, beyond the two most basic notions: category (objects together with morphisms between them, with composition of morphisms and identity morphisms) and functor (a map from objects $A$ of the source category to objects $\functor A$ of the target category, and from morphisms $f : A \to B$ to $\functor f : \functor A \to \functor B$, preserving composition and identities). In this paper, we will be working mainly with:
\begin{description}
    \item[Sets.] Objects are sets,  morphisms are functions between them.
    \item[Semigroups.] Objects are semigroups,  morphisms are semigroup homomorphisms.
\end{description}

One example of a functor is the \emph{forgetful functor} from the category of semigroups to the category of sets, which maps a semigroup to its underlying set, and a semigroup homomorphism to the corresponding function on sets. 

 \begin{example}\label{ex:functors}
    In this paper, we will mainly be studying functors from the category of semigroups to itself; such functors can be seen as semigroup constructions.  Several such constructions are listed in the following example. 
    \begin{description}
        \item[Tuples.] The functor maps a semigroup $A$ to its square $A^2$, with the semigroup operation being defined coordinate-wise. The functor extends to morphisms in the expected way. This functor also makes sense for higher powers, including infinite powers, such as $A^\omega$.
        \item[Reverse.] The functor maps a semigroup $A$ to the semigroup where the underlying set is the same, but the multiplication is reversed, i.e.~the product of $a$ and $b$ in the new semigroup is the product $ba$ in the old semigroup. Morphisms are not changed by the functor: they retain the homomorphism property despite the change in the multiplication operation.
        \item[Non-empty lists.] The functor maps a semigroup $A$ to the free semigroup $A^+$ which consists of non-empty lists (or strings) over the alphabet $A$ equipped with concatenation. On morphisms, the functor is defined element-wise (or letter-wise). 
        \item[Powerset.] The (covariant) powerset functor maps $A$ to the powerset semigroup $\powerset A$, whose underlying set is the family of all subsets of $A$, endowed with the operation
        \begin{align*}
        (A_1,A_2) \quad \mapsto \quad \left\{a_1 a_2 \mid a_1 \in A_1\ \text{and}\ a_2 \in A_2\right\}
        \end{align*}
        There are several variants of this construction: we could, for example,  require the subsets to be nonempty, or finite, or both.
        % \item Fix some semigroup $C$, and consider the functor which maps a semigroup $A$ to the sub-semigroup of the semigroup $A^C$, as in the first item, which consists only of those tuples $A^C$ that describe semigroup homomorphisms $C \to A$. For morphisms, the functor is defined coordinate-wise, as in the first item.
    \end{description}
 \end{example}

%  \begin{myexample}
%     Here is a non-example of a functor from the category of semigroups to itself. Suppose that, on objects,  we want to map each semigroup $A$ to the set of all functions $A \to A$, with the semigroup operation being function composition. The problem with this construction is that it is not clear how to extend it to morphisms, i.e.~how to map a semigroup homomorphism $f : A \to B$ to some semigroup homomorphism
%     \[
%     \begin{tikzcd}
%     (A \to A)
%     \ar[r,"\functor f"]
%     &
%     (B \to B).
%     \end{tikzcd}
%     \]
%     There are artificial ways to do this. For example, we could choose for each semigroup $B$ some distinguished element $b_0 \in B$, and map a semigroup homomorphism $f : A \to B$ to the semigroup homomorphism which maps all functions $A \to A$ to the constant function $b \mapsto b_0$. 
%  \end{myexample}
 
\noindent
 We now present the central definition of this paper. 

\newcommand{\emptytester}{2}
\begin{definition}
    A \emph{transducer semigroup} is defined to be a functor $\functor$ 
    from the category of semigroups to itself, together with an output mechanism given by a family    
    %\begin{align*}
    %\myunderbrace{\outfun_A : \functor A \to A,}{a function between two sets, that \\ is not necessarily a semigroup homomorphism}
    %\end{align*}
    \[ \outfun_A : \functor A \to A \quad\text{for each semigroup}\ A \]
    % the left alignment of the lipics style messes up the underbrace positioning
    of output functions (between sets, not necessarily semigroup homomorphisms), this family being natural in the sense that the following diagram commutes for every homomorphism $h$: 
    \[
    \begin{tikzcd}
    \functor A 
    \ar[r,"\functor h"]
    \ar[d,"\outfun_A"']
    &
    \functor B
    \ar[d,"\outfun_B"]
    \\
    A
    \ar[r,"h"']
    &
    B
    \end{tikzcd}
    \]
    A function between two semigroups $f : A \to B$, not necessarily a homomorphism, is  \emph{recognized} by a transducer semigroup if it can be decomposed as
    \[
        \begin{tikzcd}
        A 
        \ar[r,"h"]
        &
        \functor B
        \ar[r,"\outfun_B"]
        &
        B
        \end{tikzcd}
        \quad
        \text{for some semigroup homomorphism}\ h
        \]
\end{definition}
In the language of category theory, the naturality condition from the above definition says that the output mechanism is a natural transformation of type 
\[\begin{tikzcd}
    [column sep=1cm]
    {\text{Semigroups}} && {\text{Sets}}
    \arrow[""{name=0, anchor=center}, "\text{apply $\functor$ and return underlying set}", curve={height=-18pt}, from=1-1, to=1-3]
    \arrow[""{name=1, anchor=center, inner sep=0}, "\text{return underlying set (forgetful functor)}"', curve={height=18pt}, from=1-1, to=1-3]
    \arrow[ shorten <=5pt, shorten >=5pt, Rightarrow, from=0, to=1]
\end{tikzcd}\]

\begin{example}
    Consider the transducer semigroup in which the functor is the identity, and the output mechanism is also the identity. The functions of type $A \to B$ that are recognized by this transducer semigroup are exactly the semigroup homomorphisms from $A$ to $B$.
\end{example}

\begin{example}\label{ex:duplicator}
    Consider the transducer semigroup in which the functor is the identity, and the output mechanism is the function $a \in A \mapsto aa \in A$.
    %\begin{align*}
    %A & \to A\\
    %a & \mapsto aa.
    %\end{align*}

    The functions of type $A \to B$ that are recognized by this transducer semigroup are exactly those of the form $a \mapsto h(a)h(a)$ where $h$ is some homomorphism. In particular, if $h$ is the identity on $\Sigma^*$, then we get the duplicating function on strings over the alphabet $\Sigma$.
\end{example}



\begin{example}
    Consider the reversing functor from \Cref{ex:functors}. Define the output function to be the identity. Using this transducer semigroup, we can recognize the reversing function $f : \Sigma^* \to \Sigma^*$. More generally, the functions $f : A \to B$ recognized by this transducer semigroup are \enquote{anti-homomorphisms}, i.e.\ they are those that verify $f(ab) = f(b)f(a)$.
\end{example}

\begin{example}\label{ex:squaring}
    Consider the nonempty list functor $A \mapsto A^+$ described in Example~\ref{ex:functors}, with the following output mechanism:
    \begin{align*}
    A^+ & \to A\\
    [a_1,\ldots,a_n] & \mapsto \myunderbrace{(a_1 \cdots a_n) \cdots (a_1 \cdots a_n)}{$n$ times}.
    \end{align*}
    This transducer semigroup recognizes the squaring function $w \in \Sigma^+ \mapsto w^{|w|} \in \Sigma^+$ \emph{on nonempty words}, that is illustrated in the following example: $\mathtt{123 \mapsto 123123123}$.
\end{example}

\begin{example}\label{ex:squaring-empty}
    We refine the above example in order to recognize the squaring function defined on possibly empty strings: $w \in \Sigma^* \mapsto w^{|w|} \in \Sigma^*$.
    
    Let $\functor A = A^+ + A$ with the following semigroup operation: when both arguments are in~$A$ (resp.~$A^+$), we use the semigroup structure of $A$ (resp.~$A^+$); in the remaining cases, $a \cdot \ell = \ell \cdot a = \ell$ for $a \in A$ and $\ell \in A^+$. This construction on semigroups can be turned into a functor, by making the morphisms act coordinate-wise. To make it into a transducer semigroup, we define $\outfun_A : A^+ + A \to A$ by combining the output mechanism $A^+ \to A$ of the previous example with the identity $A \to A$.
    
    Let $h : \Sigma^* \to \functor(\Sigma^*)$ be the unique semigroup homomorphism such that $h(\varepsilon) = \varepsilon \in \Sigma^*$ and $h(c) = c \in (\Sigma^*)^+$ for $c\in\Sigma$. Then $\outfun_{\Sigma^*} \circ h$ is the squaring function on $\Sigma^*$, as we wanted.
\end{example}

%\begin{example}\label{ex:squaring-generalized}
%     Here is a generalization of the previous example. The functor continues to be $A^+$. The output mechanism $A^+ \to A$ is given by a sequence of strings 
%     \begin{align*}
%     w_1 \in \set{1}^+, \quad w_2 \in \set{1,2}^+, \quad w_3 \in \set{1,2,3}^+, \quad \ldots.
%     \end{align*}
%     When applied to lists of  length $n$, the output mechanism is
%     \[
%         \begin{tikzcd}
%             [column  sep=3cm]
%         A^n
%         \ar[r,"f \mapsto \text{$f$ applied to $w_n$}"]
%         &
%         A^+ 
%         \ar[r,"\text{semigroup operation}"]
%         & 
%         A.
%         \end{tikzcd}
%         \]
% \end{example}

% \begin{myexample} This example is more challenging than the previous ones, and it is meant to describe copyful \sst.
%     In this example, we use a slightly different setup: we assume that the output mechanism is a partial function, but still natural. For some finite set $R$ of \emph{register names} with a designated \emph{output register}. Define a functor where  
%     \begin{align*}
%     \functor A = R \to (R^+ \oplus A),
%     \end{align*}
%     where $\oplus$ is the co-product of semigroups, and the  semigroup structure of $\functor A$ is defined as follows. The product of two elements $f,g, \in \functor A$ is the composition of the functions described below
%     \[
%     \begin{tikzcd}
%     R
%     \ar[r,"f"]
%     &
%     R^+ \oplus A
%     \ar[r, "g^+ \oplus \id"]
%     &
%     (R^+ \oplus A)^+ \oplus A
%     \ar[r,"\text{mult} \oplus \id"]
%     &
%     R^+ \oplus A \oplus A
%     \ar[r,"\text{merge}"]
%     &
%     R^+ \oplus A.
%     \end{tikzcd}
%     \]
%     Unlike the functors in the previous examples, this functor does outputs an infinite semigroup even if the input semigroup $A$ is finite. 

%     The output mechanism is a partial function, which is obtained by selecting some fixed register. Once we have fixed the initial register, we get an output map as follows: given an element of $\functor A$, we apply it to the initial register, getting an element of $R^+ \oplus A$. If the element is in $A$, then we send it to the output; otherwise the output is undefined.

%     The functions recognized by this transducer semigroup are exactly those that can be recognized by copyful \sst with one state. 
% \end{myexample}



\subsection{Two simple characterizations}
We begin with two theorems, with simple proofs, that describe two classes of string-to-string functions in terms of the transducer semigroups that can be used to recognize them: all functions (Theorem~\ref{thm:all-functions}) and functions that reflect recognizability (Theorem~\ref{thm:reco-reflecting-functions}).
%These two theorems have simple proofs.
In the next Section~\ref{sec:reg-char} we present a third, more  interesting, theorem about  regular functions.

\subparagraph{All functions.} The first theorem shows that every function between two semigroups is recognized by some transducer semigroup.

\begin{theorem}\label{thm:all-functions} 
     Every function    between two semigroups, not necessarily a semigroup homomorphism, is recognized  by some transducer semigroup.
\end{theorem}
\begin{proof}
    Consider some semigroup $A$. Based on this semigroup we will define a transducer semigroup which will recognize all functions from $A$ to some other semigroup. Define  a functor as follows: $\functor B = A \times \displaystyle\myunderbrace{(A \to B)}{the set of all  functions, viewed as a semigroup \\ 
with the trivial semigroup operation $xy = x$}$.

The functor is defined on morphisms
as follows: the first coordinate, corresponding to~$A$, is not changed, and the second coordinate, corresponding to the set of functions, is transformed   coordinate-wise, when viewed as a tuple indexed by~$A$. (This is similar to the tuple construction in Example~\ref{ex:functors}, except that the semigroup structure of $A \to B$ is not defined coordinate-wise.)  This is easily seen to be a functor. The output mechanism, which is easily seen to be natural, is function application i.e.~$(a,f) \mapsto f(a)$.

The transducer semigroup defined above recognizes the function $f : A \to B$. The appropriate homomorphism  is~$a \in A  \mapsto  (a,f)$.
\end{proof}

\subparagraph{Recognizability reflecting functions.} Recall that a language $A \xrightarrow{\;L\;} \set{\text{yes, no}}$
%\[
%\begin{tikzcd}
%A 
%\ar[r,"L"]
%&
%\set{\text{yes, no}}
%\end{tikzcd}
%\]
is called \emph{recognizable} if it factors through a homomorphism from $A$ to some \emph{finite} semigroup.  More generally, recognizable maps from the semigroup $A$ to some set, not necessarily with two elements, are those that factor through some homomorphism into a finite semigroup.
\begin{definition}
    A function is called \emph{recognizability reflecting} when
    inverse images of recognizable languages are also recognizable.
\end{definition}
%There are a lot of recognizability reflecting functions:
\begin{example}
    Consider the semigroup $\Nat$ of natural numbers with addition. In this semigroup, the recognizable languages are ultimately periodic subsets. Consider
    \[ f : n \mapsto g(n)! \qquad\text{for any non-decreasing}\quad g : \Nat \to \Nat \]
    A property of the factorial operation $(-)!$ is that every ultimately periodic subset of $\Nat$ contains all or no factorials, up to finitely many exceptions. Therefore, the inverse image of any ultimately periodic set under $f$ will be either finite or co-finite, and therefore also ultimately periodic.
    
    Thus, there are uncountably many recognizability preserving functions of type $\Nat \to \Nat$.
\end{example}
% We now present a second characterization, which concerns functions between semigroups that are recognizability reflecting.

\begin{theorem}\label{thm:reco-reflecting-functions}
     The following conditions are equivalent for a function $f : A \to B$, which is not necessarily a semigroup homomorphism:
    \begin{enumerate}
        \item \label{it:reco-refl} $f$ is recognizability reflecting.
        %, which means for every recognizable language 
        %\[
%\begin{tikzcd}
%B 
%\ar[r,"L"]
%&
%\set{\text{yes, no}}
%\end{tikzcd}
%\]
        %the language $f;L$ is also recognizable.
        \item \label{it:trans-semig-reco}$f$ is recognized by a transducer semigroup  such that the output mechanism 
        $\outfun_B : \functor B \to B$
        is recognizable for every finite semigroup $B$.
    \end{enumerate}
\end{theorem}
\tito{\Cref{ex:squaring}?}
\begin{proof}
    For the implication (\ref{it:reco-refl}) $\Rightarrow$ (\ref{it:trans-semig-reco}) we  use a similar construction as in the proof of Theorem~\ref{thm:all-functions}.
    Consider some function $f : A \to B$, which is recognizability reflecting. Define  a functor as follows:
    $\functor C = A \times \displaystyle\myunderbrace{\mathrm{Hom}(B,C)}{the set of all semigroup homomorphisms, endowed\\
    with the trivial semigroup operation $xy = x$}$.

    On morphisms, the functor is defined as in the proof of Theorem~\ref{thm:all-functions}, and the output mechanism  is function application with $f$ inserted as an interface i.e.~$\outfun : (a,g) \mapsto g(f(a))$.

    By the assumption that $f$ is recognizability reflecting, the output mechanism $\outfun_C$ is recognizable for finite $C$. The transducer semigroup defined above recognizes the function $f$ via the homomorphism $a \in A  \mapsto  (a,\id)$.
    \tito{not sure $\outfun_C$ recognizable holds when $B$ isn't finitely generated?}
    
    
\proofsubparagraph{Consider now the converse implication (\ref{it:reco-refl}) $\Leftarrow$ (\ref{it:trans-semig-reco}).} Take a function $f : A \to B$ that satisfies~(\ref{it:trans-semig-reco}), i.e.~it is a composition $A \xrightarrow[\;h\;]{} \functor B \xrightarrow[\;\outfun_B\;]{} B$
%\[\begin{tikzcd}
%    [column sep=2cm]
%	A & {\functor B} & B 
%	\arrow["h", from=1-1, to=1-2]
%	\arrow["\outfun_{B}", from=1-2, to=1-3]
%\end{tikzcd}\]
where $h$ is some homomorphism.

We want to show that $f$ is recognizability reflecting. To prove this, let us consider some recognizable language over the output semigroup, i.e.\ for some finite semigroup $C$, a composition $B \xrightarrow[\;g\;]{} C \xrightarrow[\;\text{accepting set}\;]{} \set{\text{yes, no}}$.
%~a composition of some homomorphism from $B$ into a finite semigroup, followed by an arbitrary boolean-valued function
%\[
%\begin{tikzcd}
%    [column sep=2cm]
%B 
%\ar[r,"g"']
%&
%C
%\ar[r,"\text{accepting set}"']
%&
%\set{\text{yes,no}}
%\end{tikzcd}
%\]

We want to show that is inverse image of the language under $f$ is also recognizable. Consider the following diagram. 
\[\begin{tikzcd}
    [column sep=2cm]
	A & {\functor B} & & B \\
	& {\functor C} & & C & \set{\text{yes, no}}\\
    & & D
	\arrow["h", from=1-1, to=1-2]
	\arrow["\outfun_B", from=1-2, to=1-4]
	\arrow["{\functor g}"', from=1-2, to=2-2]
	\arrow["\outfun_C"', from=2-2, to=2-4]
	\arrow["g", from=1-4, to=2-4]
	\arrow["\text{accepting set}"', from=2-4, to=2-5]
    \arrow["\text{some homomorphism}"', from=2-2, to=3-3]
    \arrow["\text{some function}"', from=3-3, to=2-4]
\end{tikzcd}\]
The upper path from $A$ to $\set{\text{yes, no}}$ describes the inverse image under $f$. 
 The top-middle rectangle commutes by naturality of the output mechanism, and therefore the upper path describes the same function as the lower path. The existence of a bottom-middle commuting triangle, with $D$ \emph{finite}, is exactly the assumption that $\outfun_C$ is recognizable (since $C$ is finite). Finally, the lower path is a recognizable function, since the first three arrows on it are semigroup homomorphisms and $D$ is finite.
\end{proof}

% The straightforward construction in the above proof could be extended to characterization functions which reflect other properties of languages, such as being context-free or decidable. 
